Simple repression draft

### 2020-10-17 doc/resubmission/section_04_bayesian_inference.tex
minor fixes to typos

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use when stating Bayes
theorem. Again denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
m_N\}$, here Bayes' theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood -- the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom} -- so we only need to choose priors on $k_i$ and $b$.
For the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one parameter to two. Details on the arguments behind our prior
distribution selection are left for Appendix~\ref{sec:bayesian}. We state here
that the natural scale to explore these parameters is logarithmic. This is
commonly the case for parameters for which our previous knowledge based on our
domain expertise spans several orders of magnitude. For this we chose log-normal
distributions for both $k_i$ and $b$. Details on the mean and variance of these
distributions can be found in Appendix~\ref{sec:bayesian}.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
distribution function (ECDF) of the observed population distribution of mRNA
transcripts under the control of a constitutive \textit{lacUV5} promoter is
shown in black. The median posterior predictive ECDFs for models (1), Poisson,
and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is labeled with a unique number in
both (C) and (D) for cross comparison and for comparison with 
Figure~\ref{fig2:constit_cartoons}(B).}
\label{fig:constit_post_full}
\end{figure}

@@

\subsubsection{Connecting inferred parameters to prior work}
It is interesting to connect these inferences on $k_i$ and $b$ to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state thermodynamic model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m
\right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the equilibrium model ($P/N_{NS} 
\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in~\cite{Brewster2012}
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

@@

Now that we have a satisfactory model in hand for constitutive promoters, we
would like to return to the main thread: can we reconcile the thermodynamic and
kinetic models by putting to the test Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the
correspondence between indirectly inferred thermodynamic binding energies and
kinetic rates? To make this comparison, is it possible to infer repressor
binding and unbinding rates from mRNA distributions over a population of cells
as measured by single-molecule Fluorescence \textit{in situ} Hybridization
in~\cite{Jones2014}? If so, how do these inferred rates compare to direct
single-molecule measurements such as from~\cite{Hammar2014} and to binding
energies such as from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were
inferred under the assumptions of the thermodynamic models in
Figure~\ref{fig1:means_cartoons}(B)? And can this comparison shed light on the
unreasonable effectiveness of the equilibrium models, for instance, in their
application in~\cite{Chure2019, Chure2019a}?

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig04.pdf}
\caption{\textbf{Simple repression parameter inference and comparison.}
(A) Contours which enclose 50\% and 95\% of the posterior probability mass are
shown for each of several 2D slices of the 9D posterior distribution. The model
assumes one unbinding rate for each operator (Oid, O1, O2) and one binding rate
for each aTc induction concentration (corresponding to an unknown mean repressor
copy number).
(B, upper) Ratios of our inferred unbinding rates are
compared with operator binding energy differences measured by Garcia and
Phillips~\cite{Garcia2011a} (triangles) and Razo-Mejia et.\
al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
compare O1-Oid. Points with perfect agreement would lie on the dotted line.
(B, lower) Unbinding rates for O1 (cyan) and Oid (red)
inferred in this work are compared with single-molecule measurements
from Hammar et.\ al.~\cite{Hammar2014}. We plot the comparison assuming illustrative mRNA
lifetimes of $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min
(squares). Dotted line is as in upper panel.
(C) Theory-experiment comparison are shown for each of the datasets
used in the inference of the model in (A). Observed single-molecule mRNA counts
data from~\cite{Jones2014} are plotted as black lines. The median of the randomly
generated samples for each condition is plotted as a dark colored line. Lighter
colored bands enclose 95\% of all samples for a given operator/repressor copy
number pair.
%Samples are sorted by operator and plotted separately for visual clarity.
The unregulated promoter, \textit{lacUV5}, is shown with each as a reference.}
\label{fig4:repressed_post_full}
\end{figure}

@@

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
nine different conditions, spanning several combinations of three unique
repressor binding sites (the so-called $Oid$, $O1$, and $O2$ operators) and four
unique repressor copy numbers. We assume that the values of $k_i$ and $b$ are
known, since we have already cleanly inferred them from constitutive promoter
data, and further we assume that these values are the same across datasets with
different repressor binding sites and copy numbers. In other words, we assume
that the regulation of the transcription factor does not affect the mean burst
size nor the burst initiation rate. The regulation occurs as the promoter is
taken away from the transcriptionally active state when the promoter is bound by
repressor. We assume that there is one unbinding rate parameter for each
repressor binding site, and likewise one binding rate for each unique repressor
copy number. This makes our model seven dimensional, or nine if one counts $k_i$
and $b$ as well. Note that we use only a subset of the datasets from Jones et.\
al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayesian}.

@@

On the whole then, we regard this as striking confirmation of the validity of
the equilibrium models. Their lynchpin parameter is a phenomenological free
energy of repressor binding that has previously only been inferred indirectly.
Our result shows that the microscopic interpretation of this free energy, as the
log of a ratio of transition rates, does indeed hold true to within the inherent
uncertainties that remain in the entire theory-experiment dialogue.
%\mmnote{Does this deserve qualifiers, like ``to a good approximation,'' or ``as a first order model'' or something? I like the simple finality without, but I worry it's too brazen or subtlely misleading.}

@@

Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}(C) is
overall of good agreement. There some oddities, in particular the long tails in
the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since
it extends beyond the tail of the unregulated UV5 distribution. This is a
relatively small number of cells, however, so whether this is a peculiarity of
the experimental data, a statistical fluke of small numbers, or a real
biological effect is unclear. It is conceivable that there is some very slow
timescale switching dynamics that could cause this bimodality, although it is
unclear why it would only appear for specific repressor copy numbers. There is
also a small offset between experiment and simulation for O2 at the higher
repressor copy numbers, especially at 2 and 10~ng/mL. From the estimate of
repressor copy numbers from~\cite{Jones2014}, it is possible that the repressor
copy numbers here are becoming large enough to partially invalidate our
assumption of a separation of timescales between burst duration and repressor
association rate. Another possibility is that the very large number of zero mRNA
counts for Oid, 2~ng/mL is skewing its partner datasets through the shared
association rate. None of these fairly minor potential caveats cause us to
seriously doubt the overall correctness of our model, which further validates
its use to compare the equilibrium models' binding energies to the
nonequilibrium models' repressor kinetics, as we originally set out to do.

>>>>

As for the Poisson promoter model, this expression for the steady-state mRNA
distribution is exactly the likelihood we want to use when stating Bayes
theorem. Denoting the single-cell mRNA count data as $D=\{m_1, m_2,\dots,
m_N\}$, Bayes' theorem takes the form
\begin{equation}
p(k_i, b \mid D) \propto p(D\mid k_i,b)p(k_i, b).
\end{equation}
We already have our likelihood --the product of $N$ negative binomials as
Eq.~\ref{eq:neg_bionom}-- so we only need to choose priors on $k_i$ and $b$. For
the datasets from~\cite{Jones2014} that we are analyzing, as for the Poisson
promoter model above we are still data-rich so the prior's influence remains
weak, but not nearly as weak because the dimensionality of our model has
increased from one parameter to two. Details on the arguments behind our prior
distribution selection are left for Appendix~\ref{sec:bayesian}. We state here
that the natural scale to explore these parameters is logarithmic. This is
commonly the case for parameters for which our previous knowledge based on our
domain expertise spans several orders of magnitude. For this we chose log-normal
distributions for both $k_i$ and $b$. Details on the mean and variance of these
distributions can be found in Appendix~\ref{sec:bayesian}.

@@

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig03.pdf}
\caption{\textbf{Constitutive promoter posterior inference and model
comparison.} (A) The joint posterior density of model 5, the bursty promoter
with negative binomially-distributed steady state, is plotted with MCMC samples.
1D marginal probability densities are plotted as flanking histograms. The model
was fit on \textit{lacUV5} data from~\cite{Jones2014}. (B) The empirical
cumulative distribution function (ECDF) of the observed population distribution
of mRNA transcripts under the control of a constitutive \textit{lacUV5} promoter
is shown in black. The median posterior predictive ECDFs for models (1),
Poisson, and (5), negative binomial, are plotted in dark green and dark blue,
respectively. Lighter green and blue regions enclose 95\% of all posterior
predictive samples from their respective models. Model (1) is in obvious
contradiction with the data while model (5) is not. Single-cell mRNA count data
is again from~\cite{Jones2014}. (C) Joint posterior distributions for burst rate
$k_i$ and mean burst size $b$ for 18 unregulated promoters
from~\cite{Jones2014}. Each contour indicates the 95\% highest posterior
probability density region for a particular promoter. Note that the vertical
axis is shared with (D). (D) Plots of the burst rate $k_i$ vs.\ the binding
energy for each promoter as predicted in~\cite{Brewster2012}. The dotted line
shows the predicted slope according to Eq.~\ref{eq:bursty_equil_corresp1},
described in text. Each individual promoter is labeled with a unique number in
both (C) and (D) for cross comparison and for comparison with
Figure~\ref{fig2:constit_cartoons}(B).}
\label{fig:constit_post_full}
\end{figure}

@@

\subsubsection{Connecting inferred parameters to prior work}
It is interesting to connect these inferences on $k_i$ and $b$ to the work
of~\cite{Brewster2012}, where these same 18 promoters were considered through
the lens of the three-state thermodynamic model (model 2 in
Figure~\ref{fig1:means_cartoons}(B)) and binding energies $\Delta\varepsilon_P$
were predicted from an energy matrix model derived from~\cite{Kinney2010}. As
previously discussed the thermodynamic models of gene regulation can only make
statements about the mean gene expression. This implies that we can draw the
connection between both frameworks by equating the mean mRNA $\left\langle m
\right\rangle$. This results in
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma}
        \frac{\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}
                {1+\frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P)}.
\end{equation}
By taking the weak promoter approximation for the thermodynamic model ($P/N_{NS}
\exp(-\beta\Delta\varepsilon_r) \ll 1$) results in~\cite{Brewster2012}
\begin{equation}
\langle m \rangle = \frac{k_i b}{\gamma}
        = \frac{r}{\gamma} \frac{P}{N_{NS}}\exp(-\beta\Delta\varepsilon_P),
\end{equation}
valid for all the binding energies considered here.

@@

Now that we have a satisfactory model in hand for constitutive promoters, we
would like to return to the main thread: can we reconcile the thermodynamic and
kinetic models by putting to the test Eq.~\ref{eq:deltaFR_eq_noneq_equiv}, the
correspondence between indirectly inferred thermodynamic binding energies and
kinetic rates? To make this comparison, is it possible to infer repressor
binding and unbinding rates from mRNA distributions over a population of cells
as measured by single-molecule Fluorescence \textit{in situ} Hybridization
in~\cite{Jones2014}? If so, how do these inferred rates compare to direct
single-molecule measurements such as from~\cite{Hammar2014} and to binding
energies such as from~\cite{Garcia2011a} and~\cite{Razo-Mejia2018}, which were
inferred under the assumptions of the thermodynamic models in
Figure~\ref{fig1:means_cartoons}(B)? And can this comparison shed light on the
unreasonable effectiveness of the thermodynamic models, for instance, in their
application in~\cite{Chure2019, Chure2019a}?

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig04.pdf}
\caption{\textbf{Simple repression parameter inference and comparison.}
(A) Contours which enclose 50\% and 95\% of the posterior probability mass are
shown for each of several 2D slices of the 9D posterior distribution. The model
assumes one unbinding rate for each operator (Oid, O1, O2) and one binding rate
for each aTc induction concentration (corresponding to an unknown mean repressor
copy number). (B, upper) Ratios of our inferred unbinding rates are compared
with operator binding energy differences measured by Garcia and
Phillips~\cite{Garcia2011a} (triangles) and Razo-Mejia et.\
al.~\cite{Razo-Mejia2018} (circles). Blue glyphs compare O2-O1, while orange
compare O1-Oid. Points with perfect agreement would lie on the dotted line. (B,
lower) Unbinding rates for O1 (green) and Oid (red) inferred in this work are
compared with single-molecule measurements from Hammar et.\
al.~\cite{Hammar2014}. We plot the comparison assuming illustrative mRNA
lifetimes of $\gamma^{-1}=3$~min (triangles) or $\gamma^{-1}=5$~min (squares).
Dotted line is as in upper panel. (C) Theory-experiment comparison are shown for
each of the datasets used in the inference of the model in (A). Observed
single-molecule mRNA counts data from~\cite{Jones2014} are plotted as black
lines. The median of the randomly generated samples for each condition is
plotted as a dark colored line. Lighter colored bands enclose 95\% of all
samples for a given operator/repressor copy number pair. The unregulated
promoter, \textit{lacUV5}, is shown with each as a reference.}
\label{fig4:repressed_post_full}
\end{figure}

@@

Before showing the mathematical formulation of our statistical inference model,
we would like to sketch the intuitive structure. The dataset
from~\cite{Jones2014} we consider consists of single-cell mRNA counts data of
nine different conditions, spanning several combinations of three unique
repressor binding sites (the so-called Oid, O1, and O2 operators) and four
unique repressor copy numbers. We assume that the values of $k_i$ and $b$ are
known, since we have already cleanly inferred them from constitutive promoter
data, and further we assume that these values are the same across datasets with
different repressor binding sites and copy numbers. In other words, we assume
that the regulation of the transcription factor does not affect the mean burst
size nor the burst initiation rate. The regulation occurs as the promoter is
taken away from the transcriptionally active state when the promoter is bound by
repressor. We assume that there is one unbinding rate parameter for each
repressor binding site, and likewise one binding rate for each unique repressor
copy number. This makes our model seven dimensional, or nine if one counts $k_i$
and $b$ as well. Note that we use only a subset of the datasets from Jones et.\
al.~\cite{Jones2014}, as discussed more in Appendix~\ref{sec:bayesian}.

@@

On the whole then, we regard this as striking confirmation of the validity of
the thermodynamic models. Their lynchpin parameter is a phenomenological free
energy of repressor binding that has previously only been inferred indirectly.
Our result shows that the microscopic interpretation of this free energy, as the
log of a ratio of transition rates, does indeed hold true to within the inherent
uncertainties that remain in the entire theory-experiment dialogue.

@@

Our broad brush takeaway from Figure~\ref{fig4:repressed_post_full}(C) is
overall of good agreement. There some oddities, in particular the long tails in
the data for Oid, 1~ng/mL, and O2, 0.5~ng/mL. The latter is especially odd since
it extends beyond the tail of the unregulated UV5 distribution. This is a
relatively small number of cells, however, so whether this is a peculiarity of
the experimental data, a statistical fluke of small numbers, or a real
biological effect is unclear. It is conceivable that there is some very slow
timescale switching dynamics that could cause this bimodality, although it is
unclear why it would only appear for specific repressor copy numbers. There is
also a small offset between experiment and simulation for O2 at the higher
repressor copy numbers, especially at 2 and 10~ng/mL. From the estimate of
repressor copy numbers from~\cite{Jones2014}, it is possible that the repressor
copy numbers here are becoming large enough to partially invalidate our
assumption of a separation of timescales between burst duration and repressor
association rate. Another possibility is that the very large number of zero mRNA
counts for Oid, 2~ng/mL is skewing its partner datasets through the shared
association rate. None of these fairly minor potential caveats cause us to
seriously doubt the overall correctness of our model, which further validates
its use to compare the thermodynamic models' binding energies to the kinetic
models' repressor binding and unbinding rates, as we originally set out to do.
